---
title: "Hospital Readmission Risk Stratification: Identifying High-Risk Diabetic Patients for Proactive Clinical Intervention"
output: html_document
date: "2025-02-09"
author: Nusrath Ameena Ahamed Nasar 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction 

Context: 
An end-to-end predictive analytics system to identify diabetic patients at high risk of 30-day readmission, enabling hospitals to prioritize discharge planning and post-acute interventions.

Question: 
1. Which diabetic patients are at highest risk of 30-day hospital readmission
2. what clinical and operational factors drive this risk
3. How can hospitals optimize discharge planning and post-acute care to reduce preventable readmissions?

### Read in data 
```{r}
df = read.csv('data/dataset.csv', header = TRUE)
num_cols <- unlist(lapply(df, is.numeric))
plot(df[,num_cols])
```

```{r}
colnames(df)
```


```{r}
df$age <- as.numeric(df$age)
df$children <- as.numeric(df$children)

for (col in colnames(df)) {
  cat("Column:", col, "- Type:", class(df[[col]]), "\n")
}
```



```{r}
# Loop through columns and create visualizations with custom axis labels
for (col in colnames(df)) {
  if (is.numeric(df[[col]])) {
    hist(df[[col]], 
         main = paste("Histogram of", col), 
         xlab = paste(col, "Values"),  # Custom x-axis label
         ylab = "Frequency",           # Custom y-axis label
         col = "skyblue")
  } else if (is.character(df[[col]]) || is.factor(df[[col]])) {
    barplot(table(df[[col]]), 
            main = paste("Bar Chart of", col), 
            xlab = col,                # Custom x-axis label
            ylab = "Count",            # Custom y-axis label
            col = "pink")
  }
}

```

```{r}
# Create a frequency table for a specific column
freq_table <- table(df$sex)

# Create the bar chart
bar_positions <- barplot(freq_table, 
                         main = "Bar Chart with Counts", 
                         xlab = "Categories", 
                         ylab = "Frequency", 
                         col = "skyblue")

# Add counts on top of bars
text(x = bar_positions, 
     y = freq_table, 
     labels = freq_table, 
     pos = 1)

```
```{r}
freq_table <- table(df$smoker)

# Create the bar chart
bar_positions <- barplot(freq_table, 
                         main = "Smoker Chart with Counts", 
                         xlab = "Categories", 
                         ylab = "Frequency", 
                         col = "skyblue")

# Add counts on top of bars
text(x = bar_positions, 
     y = freq_table, 
     labels = freq_table, 
     pos = 1)

```


```{r}
library(ggplot2)

# Replace "column_name" with your numeric column
ggplot(df, aes(x = medical.charges)) +
  geom_density(aes(y = after_stat(count)),fill = "lightblue", color = "blue") +
  labs(title = "Medical Charges", x = "Amount", y = "Density")

```
Correlation between cost to variables 
a. Numeric Data type 
```{r}

cor(df[sapply(df, is.numeric)], method = "spearman")

```
b. Character data type
```{r}
smoker = as.factor(df$smoker)
sex = as.factor(df$sex)
region = as.factor(df$region)

boxplot(df$medical.charges ~ smoker, main = 'Smoker')
boxplot(df$medical.charges ~ sex, main = 'Sex')
boxplot(df$medical.charges ~ region, region = 'Region')
```


Next steps:

- region, maybe also include a map - perhaps we can combine with powerbi?
